---
title: "Homework-1"
author: "Alice Kemp"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
if (!("librarian" %in% rownames(utils::installed.packages()))) {
  utils::install.packages("librarian")
}
librarian::shelf(
  ggthemes,
  tidyverse,
  dplyr,
  haven, 
  stargazer, 
  RColorBrewer,
  here,
  ggmap,
  scales, 
  ggridges,
  viridis,
  gganimate, 
  gifski, 
  future, 
  geosphere,
  maps, 
  airportr, 
  kable, 
  kableExtra,
  rsample,
  caret,
  modelr,
  parallel,
  foreach,
  scales, 
  transformr)

my_theme = theme_minimal(base_family = "Arial Narrow", base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = rel(1.7)),
        plot.subtitle = element_text(face = "plain", size = rel(1.3), color = "grey70"),
        plot.caption = element_text(face = "italic", size = rel(0.7), 
                                    color = "grey70", hjust = 0),
        legend.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold", size = rel(1.1), hjust = 0),
        axis.title = element_text(face = "bold"),
        axis.title.x = element_text(margin = margin(t = 10), hjust = 0),
        axis.title.y = element_text(margin = margin(r = 10), hjust = 1))
```

## **1) Data visualization: flights at ABIA**


```{r abia, echo = FALSE}
ABIA = read_csv(here("data/ABIA.csv"), show_col_types = FALSE) %>%
  select(DayOfWeek, UniqueCarrier, ArrDelay, DepDelay, Origin, Dest) %>%
  filter(Origin == "AUS") %>%
  mutate(delay_total = ArrDelay - DepDelay,
         DayOfWeek = recode(DayOfWeek, 
                            "1"="Monday",
                            "2"="Tuesday",
                            "3"="Wednesday",
                            "4"="Thursday",
                            "5"="Friday",
                            "6"="Saturday",
                            "7"="Sunday"),
         DayOfWeek = fct_relevel(DayOfWeek, 
                                 "Monday", "Tuesday", "Wednesday", 
                                 "Thursday", "Friday", "Saturday", 
                                 "Sunday"))
attach(ABIA)

gganimate = ggplot(ABIA, aes(x=delay_total, y=UniqueCarrier, fill = stat(x))) +
  geom_density_ridges_gradient(rel_min_height = 0.01, scale = 2, show.legend = FALSE) +
  scale_fill_viridis(option = "plasma", alpha = 0.9, discrete = FALSE, direction= -1) + 
  xlim(-40,100) + 
  transition_states(DayOfWeek, transition_length = 1, state_length = 2) + 
  labs(
    title  = "ABIA Outbound Flights: Net delay by carrier",
    subtitle = "Day of Week: {closest_state}",
    caption = "The most consistent airlines are ExpressJet (XE), Southwest Airlines (WN), and American Airlines (AA), while net delays for \n Atlantic Southeast Airlines (EV) and Northwest Airlines (NW) tend to vary more between days.") + 
  xlab("Net Delay (min)") + 
  ylab("Carrier") + 
  my_theme
animate(gganimate)
```

## **2) Wrangling the Billboard Top 100**
### **Part A: Ten most popular songs since 1958**
```{r billboard_a, echo = FALSE, message = FALSE}
# clean data and filter 
billboard = read_csv(here(("data/billboard.csv")),show_col_types = FALSE) %>%
  select(performer, song, year, week, week_position) %>%
write_csv(here("data/billboard_clean.csv")) %>%
  group_by(performer, song) %>%
  filter(year >= 1958) %>%
  summarise(count = n()) %>%
  arrange(desc(count), .by_group = FALSE) 
# table of top ten songs
n = seq(1,10,1)
data.frame(n,billboard[1:10,]) 
```

### **Part B: Musical diversity of Top 100 over time**
```{r billboard_b, echo=FALSE, warning=FALSE}
billboard_unique = read_csv(here("data/billboard_clean.csv"),show_col_types = FALSE) %>%
  group_by(performer, song, year) %>%
  filter(year != 1958, year != 2021) %>%
  unite("performer_song", performer:song, sep = " - ", remove = TRUE) %>%
  summarize(
            unique_songs = n_distinct(performer_song)) 

ggplot(billboard_unique, aes(x=year, y=unique_songs)) + 
  geom_line(color = "mediumblue", alpha = 0.6, size = 1.0) + 
  ylab("unique songs") + 
  ggtitle('Unique Billboard Top 100 songs by year') + 
  labs(caption = 'The number of unique songs on the Billboard Top 100 list peaked in the late 1960s, before dipping to a low in 2001. \n However, music diversity has risen again, with 2020 reaching 800 unique songs on the Top 100.') + 
  my_theme + 
  theme(
    plot.caption = element_text(hjust = 0))
```


### **Part C: Artists with ten-week hits**
```{r billboard_c, echo=FALSE, message = FALSE}
read_csv(here("data/billboard_clean.csv"),show_col_types = FALSE) %>%
  group_by(performer, song) %>%
  summarize(
    nweeks = n()) %>%
  filter(
    nweeks>=10,
    ) %>%
  group_by(performer) %>%
  summarize(nsongs = n()) %>%
  filter(nsongs >= 30) %>%
ggplot(aes(x=reorder(performer, +nsongs), y=nsongs)) + 
  geom_col(fill = "orangered2", alpha = 0.9) + 
  ylab("Ten Week Songs (>30)") + 
  xlab("") + 
  ggtitle('Billboard 100 Ten-Week Artists') + 
  labs(caption = "There have been 19 artists with 30 or more songs on the Billboard 100 for ten or more weeks since \n 1958. Elton John leads the list with 52 songs followed by Madonna (44) and Kenny Chesney (42)") + 
  my_theme + 
  theme(
    plot.caption = element_text(hjust = 0)) + 
  coord_flip() 
```


## **3) Wrangling the Olympics**

### **Part A**
The 95th percentile of female Olympic medalists in the top 20 sports between 1924 and 2016 is 186 cm. 
|                    	| **95th Percentile** 	|
|--------------------	|---------------------	|
| Female Height (cm) 	| 186.0               	|
```{r olympics_a, echo=FALSE}
olympics_female = read_csv(here("data/olympics_clean.csv"), show_col_types = FALSE) %>%
  select(id, sex, height, event, year) %>%
  filter(sex=="F") %>%
  group_by(event)
perc95 = quantile(olympics_female$height, 0.95)
attach(olympics_female)
```

### **Part B** 
The greatest variation in female medalists' height occurs in Women's Coxed Fours Rowing, where the average height is 173.3 cm with a standard deviation of 10.9 cm. Overall, Coxed Rowing events account for three out of the top five for most varied height distributions - this is due to the historically dramatic difference in optimal heights for crew rowers versus coxswains. 

|                                       	| **Average Height (cm)** 	| **Standard Deviation (cm)** 	|
|---------------------------------------	|:-----------------------:	|:---------------------------:	|
| Rowing Women's Coxed Fours            	|          173.3          	|             10.9            	|
| Women's Basketball                    	|          183.0          	|             9.7             	|
| Rowing Women's Coxed Quadruple Sculls 	|          171.9          	|             9.2             	|
| Rowing Women's Coxed Eights           	|          178.2          	|             8.7             	|
| Swimming Women's 100 metres Butterfly 	|          173.2          	|             8.1             	|
```{r olympics_b, echo = FALSE}
olympics_female %>%
  group_by(event) %>%
  summarize(
    avg = mean(height),
    sd = sd(height)) %>%
  arrange(desc(sd)) %>%
  top_n(5)
```
### **Part C**

```{r olympics_c, echo = FALSE}
# average age across genders
swim_avgage_df = data.frame(read_csv(here("data/olympics_clean.csv"), show_col_types = FALSE) %>%
  filter(sport == "Swimming") %>%
  group_by(year) %>%
  summarize(avg_age = mean(age)) %>%
  mutate(across(where(is.numeric), ~ round(., 1))))
swim_avgage_df
attach(swim_avgage_df)

# split by gender
olympics_ageyr = read_csv(here("data/olympics_clean.csv"),show_col_types = FALSE) %>%
  filter(sport == "Swimming") %>%
  group_by(year, sex) %>%
  summarize(avg_age = mean(age))
attach(olympics_ageyr)
ggplot(data = olympics_ageyr, aes(x=year,y=avg_age, color = sex)) + 
  geom_line(alpha = 0.7) +
  geom_line(data = swim_avgage_df, aes(x=year,y=avg_age, color = "darkgrey"), linetype = "dotted", inherit.aes=FALSE) + 
  scale_color_manual(values = c("M" = "mediumblue",
                                "F"="violetred2")) + 
  labs(title = "Average Age of Olympic Swimmers over Time",
       caption = "After peaking in 1924, the average age of olympics swimmers has increased over time with female swimmers \n being youngswer than male swimmers on average. However, the difference in average age for males versus \n females has decreased over time, most notably equaling zero in the year 2000.") +
  ylab("average age") + 
  my_theme + 
  theme(
    plot.caption = element_text(hjust = 0))
```


## **4) K-nearest neighbors**
The 350 trim level yields a higher optimal value of k, equal to 32. This may be due to the fact that there is less variance in the data for 350 trim models than for 65 AMG models. Thus, the 350 KNN model can optimize with a higher value of k (more neighbors, but lower sd) to minimize error as compared to the KNN model for the 65 AMG cars.  
```{r sclass}
# Trim level: 350 
sclass = read.csv("/Users/alicekemp/statistical-learning/stat-learning/Homework-1/data/sclass.csv") %>%
  select(trim, mileage, price) %>%
  filter(trim %in% c(350,"65 AMG"))

sclass350 = sclass %>%
  filter(trim==350)
attach(sclass)

# create train/test splits
sclass350_split = initial_split(sclass350, prop = 0.8)
sclass350_train = training(sclass350_split)
sclass350_test = testing(sclass350_split)

# KNN models
rmse_out350 = foreach(i=2:100, .combine='c') %do% {
  # train the model and calculate RMSE on the test set
  knn_model = knnreg(price ~ mileage, data=sclass350_train, k = i)
  modelr::rmse(knn_model, sclass350_test)
} 
##### 
# repeat for 65 AMG trim level 
#####
sclass65 = sclass %>%
  filter(trim=="65 AMG")

sclass65_split = initial_split(sclass65, prop = 0.8)
sclass65_train = training(sclass65_split)
sclass65_test = testing(sclass65_split)

rmse_out65 = foreach(i=2:100, .combine='c') %do% {
  knn_model = knnreg(price ~ mileage, data=sclass65_train, k = i)
  modelr::rmse(knn_model, sclass65_test)} 

# plot K vs. RMSE for both trim levels
k_vals = seq(2,100,1)
df = data.frame(k_vals, rmse_out350, rmse_out65) 
ggplot(df) + 
  geom_line(aes(x = k_vals, y = rmse_out350, color = "350"), size = 1.5, alpha = 0.8) + 
  geom_line(aes(x = k_vals, y = rmse_out65, color = "65 AMG"), size = 1.5, alpha = 0.8) + 
  scale_color_manual("",
                     breaks = c("350", "65 AMG"),
                     values = c("darkorange2","violetred2")) + 
  my_theme + 
  xlab("K") + 
  ylab("RMSE") + 
  labs(title = "KNN Model: K vs. RMSE") 

#####
# Trim level: 350
#####

# find optimal K (minimum RMSE)
optimal_k = df %>%
  select(k_vals, rmse_out350) %>%
  arrange(rmse_out350)

# attach predictions to the test df
knn32 = knnreg(price ~ mileage, data=sclass350_train, k = 32)

sclass350_test = sclass350_test %>%
  mutate(price_pred = predict(knn32, sclass350_test))

# plot data and KNN model fit 
ggplot(data = sclass350_test) + 
  geom_point(mapping = aes(x =  mileage, y = price), alpha=0.2) + 
  geom_line(aes(x = mileage, y = price_pred), color='darkorange2', size=1.5) +
  my_theme + 
  labs(
    title = "Mercedes S Class: Mileage vs. Price",
    subtitle = "Trim level: 350 | KNN Model (K=32)")

#####
# Trim level: 65 AMG
#####
# find optimal K (minimum RMSE) 
optimal_k = df %>%
  group_by(k_vals) %>%
  arrange(rmse_out65) %>%
  top_n(1)
# attach predictions to the test df
knn10 = knnreg(price ~ mileage, data=sclass65_train, k = 10)

sclass65_test = sclass65_test %>%
  mutate(price_pred = predict(knn10, sclass65_test))

ggplot(data = sclass65_test) + 
  geom_point(mapping = aes(x =  mileage, y = price), alpha=0.2) + 
  geom_line(aes(x = mileage, y = price_pred), color='violetred2', size=1.5) +
  my_theme + 
  labs(
    title = "Mercedes S Class: Mileage vs. Price",
    subtitle = "Trim level: 65 AMG | KNN Model (K=10)") 

```







